Density-dependent selection with strictly relative contests

### 2017-11-18 manuscript.tex
l

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are made sufficiently large. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the shape of these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

Here we compare MacArthur's claims about selection in crowded populations with our density-dependent lottery model.

In the Introduction we gave the essential details of MacArthur's general argument, but we did not specify the role of ``$K$''.

revolving around the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ of Eq.~\eqref{eq:macgeneral}

The symbol ``$K$'' enters the argument as the general label for the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.

We are now in a position to evaluate MacArthur's claim that ``fitness is $K$'' in crowded environments. Importantly, only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types. For instance, in the case of the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects between individuals of the first type, whereas $\alpha_{12}=1/K_{12}$ measures the effects of competition from individuals of the second type.  Thus, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

>>>>

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are made sufficiently large. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

We now compare MacArthur's claims about selection in crowded environments with our density-dependent lottery model.

As shown in the Introduction, MacArthur's argument revolves around relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ of Eq.~\eqref{eq:macgeneral}. To formalize this relationship, the familiar symbol ``$K$'' is introduced to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. It is in this sense that MacArthur concludes that ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}.

Notice that this use of ``$K$'' is considerably more general than the $K$ parameter in the logistic model, including both the saturation densities $K_{11}$ and $K_{22}$ and high-density interaction terms $K_{12}$ and $K_{21}$. For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second.  Thus, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
